Interactive Curvature Tensor Visualization on Digital Surfaces\thanks{This work has been mainly funded by \textsc{DigitalSnow} ANR-11-BS02-009, \textsc{KIDICO} ANR-2010-BLAN-0205 and \textsc{PRIMES Labex} ANR-11-LABX-0063/ ANR-11-IDEX-0007 research grants.}

### 2016-01-24 latex/article.tex
minor fix

Curvature estimation on discrete or digital surface has also been
widely investigated. In \cite{CVIU2014}, authors propose digital
versions of Integral Invariant estimators
\cite{Pottmann2007,Pottmann2009} in order to estimate the complete
curvature tensor (mean/Gaussian curvatures, principal curvatures,
principal curvature directions, normal vector field) on digital
surfaces. Such approaches are based on an integration principle using
a ball kernel of a given radius. Additionally, authors have
demonstrated that these estimators have multigrid convergence
properties. In this article, we also propose an efficient GPU
implementation of such estimators to visualize
%, in real-time,
such curvature fields in real-time and
%allowing us
to interactively change the
%explore different
value of the kernel radius.

\vspace{0.2cm} \txtblue{In this paper, we first present the
previous works on curvature tensor estimation. Then, we propose an
efficient GPU approach to extract a triangulated isosurface from a
digital object. Finally, we propose a fully-parallel  GPU pipeline to compute
curvature tensor in real time and present our results.}

@@

\sloppy In our context, we consider digital shapes (any subset $Z$ of $\Z^d$)
and boundaries of digital shapes $\DigBoundary{Z}$. We denote by $\DSh$ the
Gauss digitization of $\Shape$ in a $d-$dimensional grid with grid step $h$,
\emph{i.e.}  $\DSh \EqDef \{ \vz \in \Z^d,\, h\cdot\vz\in\Shape\}$. For such
digitized set $\DigShape\EqDef\DSh$, $\Body{\DigShape}{h}$ is a subset of $\R^d$
corresponding to the union of hypercubes centered at $h\cdot \vz$ for $\vz\in Z$
with edge length $h$. By doing so, both $\partial\Shape$ and
$\partial\Body{Z}{h}$ are topological boundaries of objects lying in the same
space (see Fig. \ref{fig:notations}-b). Note that the combinatorial digital
boundary $\DigBoundary{Z}$ of $Z$ made of cells in a cellular Cartesian complex
(\emph{pointels}, \emph{linels}, \emph{surfels}, \ldots), can be trivially
embedded into $\R^d$ such that it coincides with $\partial\Body{Z}{h}$.

\begin{figure}[t]{\small
    \begin{center}
      {\begin{overpic}[width=5cm]{figs/integral2D_2}
          \put(42,28){$\times$}
          \put(46,32){$\vx$}
          \put(55,45){$\Ball{R}{\vx}$}
          \put(1,10){$\dS$}
          \put(15,5){$\Shape$}
      \end{overpic}}~~~~~~
      {\begin{overpic}[width=5cm]{figs/notation_2}
          \put(62,0.5){$h \cdot \vz \in \Shape \cap \Ball{R}{\vx}$}
          \put(55,45){$\Ball{R}{\vx}$}
          \put(42,28.5){$\times$}
          \put(45,32.5){$\vx$}
          \put(45,16.5){$h$}
          %\put(88,39){$\Proj(p)$}
          \put(1.5,10){$\partial \Shape$}
          \put(15,7){$\partial \Body{\DSh}{h}$}
      \end{overpic}}
      \vspace{-0.4cm}
    \end{center}}
    \caption{Integral invariant computation (\emph{left}) and
      notations (\emph{right}) in dimension 2 \cite{CVIU2014}.\label{fig:notations}}
    \vspace{-0.35cm}
\end{figure}

In \cite{DGCI2013}, estimators are defined based on the 2D digital curvature estimator
$\CurvH{R}$ and the 3D digital mean curvature estimator $\MeanCurvH{R}$ based on
the digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ (we denote such
quantity $\AreaC(Y,h)$ in dimension 2):
%% Redite
%% \Jaco{Given some point $\vx \in \R^d$, the set $Y$ is the digital intersection of $\DigShape$ with the ball of radius $R/h$ and center $\vx/h$.}
%% where $Y$ is the intersection of $\DigShape$ with a ball
%% $\Ball{R/h}{\vx/h}$ with $\MCard(Y)$, the number of digital points of
%% the intersection between a ball $\Ball{R/h}{\vx/h}$ of digital radius
%% $R/h$ and $\DigShape$ (we denote such quantity $\AreaC(Y,h)$ in
%% dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), {\em digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
Estimators of both 2D curvature and 3D mean curvature on digital
objects with volumetric integral information are proved to be multigrid convergent
property \cite{DGCI2013}. For short, when the digital object becomes finer and
finer, \ie when the digitization step $h$ tends to zero, the estimated quantity
on $\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
\txtblue{
associated one on $\Bd{\Shape}$ in $O\left(h^{\frac{1}{3}}\right)$ for
convex shapes with at least $C^3$-boundary and bounded curvature
(setting $R\EqDef kh^\frac{1}{3}$ for some $k\in\R$).}
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
%Setting $R = kh^\frac{1}{3}$ on convex shapes
%with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
%related to the maximal curvature of the shape, they have:
%%
%\begin{align}
%\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
%\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
%&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
%&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
%\end{align}

In \cite{CVIU2014}, authors have also defined 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  {\em 3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of the intersection between the digital shape
  $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$, order $1$
and order $2$ (see Equation~20 of \cite{CVIU2014} for more details).
% ($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
% $\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
% $\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
% Equation~20 of \cite{CVIU2014} for more details).
\txtblue{
These estimators are proven convergent in $O\left(h^{\frac{1}{3}}\right)$ when setting the ball
radius $h$ in $R = kh^\frac{1}{3}$, where $k$ is a constant related to the
maximal curvature of the shape, on convex shapes with at least $C^3$-boundary
and bounded curvature \cite{CVIU2014}.}
%More formally, it leads to:
%%
%\begin{align}
%  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
%  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
%  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
%  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
%\end{align}

Additionally, eigenvectors associated to $\hat{\lambda}_1$ and
$\hat{\lambda}_2$ of the covariance matrix are principal
curvature direction estimators $\PrincDirH{1}{R}$ and
$\PrincDirH{2}{R}$. The smallest eigenvector corresponds to the normal
direction $\NormalDirH{R}$ at $\vx$. Convergence results can be found
in \cite{ChapterIICurvature}.
%%% Changer avec le chapitre de Jaco à un moment

It has been shown that the radius of the ball depends on the geometry of the
underlying shape. In \cite{DGCI2014}, a proposal was made for a parameter-free
estimation of the radius of the ball by analyzing the shape \wrt the local shape
geometry using maximal digital straight segments of the digital boundary. In
\cite{SMI2015}, these estimators have been analyzed in scale-space (for a range of radii)  for a given digital shape. This allows to detect features of the
shape thanks to the behavior of estimators on singularities.

As a consequence, for all these integral invariant based approaches, we need to
consider different ball radius which could be time consuming when implemented on
CPU. We propose here a fully parallel implementation on GPU allowing us to
change the radius and thus update the estimated quantities in real-time.

@@

Using this spatial data structure, a triangulated mesh can be constructed
%from a Marching Cubes approach \cite{lorensen1987marching} (MC for short): the
using Marching Cubes \cite{lorensen1987marching} (MC for short): the
triangulation is generated from local triangle patches computed on
local cell configurations.
%Since triangle patches are locally computed,
Such approach is fully parallel and easy to implement on the GPU.
However, since adjacent cells may not have the same depth
in the octree, original \textsc{Lorensen} and \textsc{Cline}'s rules need to be
updated (see Figure~\ref{fig_quadtree_partitionning}-\emph{right}).
Many authors have addressed this problem both for primal and
dual meshes
\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.

@@

As illustrated in Figure~\ref{fig_quadtree_partitionning}-\emph{middle}, we
%propose a viewpoint dependent criterion to decide if a cell needs to be refined:
propose a viewpoint dependent criterion to decide if a cell needs to be refined:
the closer we are to the camera, the finer the cells are.
%In addition to the
%distance criterion, we define an angular criterion to prevent refining cells
%outside the camera frustrum.
Such a criterion is also well suited to GPU since it can be evaluated
independently on every cell.
%Hence, we can efficiently determine if a cell must be subdivided or not.
Figure~\ref{fig_lod_octree} illustrates our level of
details (LoD for short) criterion.
% we use. More precisely,
In dimension 2, if $\alpha$ denotes the viewing angle, an object at a distance $d$ from the camera
%with diameter greater than $2\cdot d\cdot\tan(\alpha)$ is projected into more
has a projected size on screen of $2\cdot d\cdot\tan(\alpha)$.
%than one pixel
(see Figure~\ref{fig_lod_octree}-\emph{left}).
%Hence, the
Our distance criterion is based on the ratio (\emph{visibility ratio} in the
following) between the cell diameter $l(c)$ (power of 2 depending on the depth),
and its projected size. %$2\cdot d_c\cdot\tan(\alpha)$, $d_c$ being the distance between the cell
%center and the camera.
For a given cell $c$, split and merge decision are based
on this visibility ratio:
%
\begin{itemize}
  \item $c$ is split if its children
cells have a visibility ratio greater than  constant $k$\,;
  \item $c$ and its sibling cells are merged if their parent cell $c'$ has a visibility ratio lower than $k$\;
  \item otherwise, the cell $c$ stays for the next frame.
\end{itemize}
%
\begin{figure}[!h]
  \begin{center}
    \begin{overpic}[width=3.5cm]{figs/criterion}
      \put(33,40){$\alpha$}
      \put(50,45){$d_c$}
      \put(95,65){$l(c)$}
    \end{overpic}~~~~~
    \includegraphics[width=3cm]{viewlod2_small}~~~~~
    %\includegraphics[width=3cm]{subdivision}
  \end{center}
  \caption{Notations \emph{(left)} and adaptive meshing in dimension 2 using the LoD distance and angular criterion \emph{(right)}.
  %Based on LoD criterion evaluation, the cell buffer is updated between each frame \emph{(right)}.
  }
  \label{fig_lod_octree}
\end{figure}
%
%In addition to the distance criterion, we add an angular term in order
%to prevent subdivision of cells far from the camera frustrum (see
%Figure~\ref{fig_lod_octree}-\emph{middle}). We do not go into the
%details but please note that stay, split and merge decisions of all cells are computed in
%parallel from all cell morton codes.
Using such a criterion, split and merge decisions are computed in
%of all cells are computed in
parallel from the morton codes of all cells.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{figs/pipeline2}
  \caption{GPU pipeline summary. Data buffers are
    represented in green and computations in red. Each computation
    retrieves data from a buffer and fills a new one. }
  \label{fig_pipeline}
\end{figure}

@@

\paragraph{Naive Approach.}
\txtblue{A first simple solution  consists in scanning all digital
points, at a given resolution, lying inside the integration domain}
(Fig.~\ref{fig:approx}-$left$) and then \txtblue{estimating} the
geometrical moment as the sum of the geometrical moments of each elementary
cubes lying in the intersection (see
Eq.~(\ref{eq:dig-curvature-estimator-k}) to (\ref{eq:dig-curvature-estimator-k2})).
\txtblue{
This is exactly similar to what is done on the CPU. On  the GPU, we can
exploit \emph{mipmap} textures to obtain  multi-resolution information.}
If the input binary object is stored in a
3D texture, GPU hardware constructs a multi-resolution pyramid (mipmap) such that 8
neighboring voxel intensities at level $l$ are averaged to define the voxel
value at level $l+1$. If the level 0 corresponds to the binary input object, at
\txtblue{
a given level $l$, a texture probe at  a point $(x,y,z)$ returns the fraction of $\DSh$ belonging}
to the cube of center $(x,y,z)$ and edge length $2^l$. As a consequence, we can
approximate the volume of $\Ball{R}{\vx}\cap\Shape$ by considering mipmap values
at a given resolution $l$ (Fig.~\ref{fig:approx}-$right$). In this case, errors only
occurs for cells lying inside $\Shape$ (with density 1) not entirely covered by
$\Ball{R}{\vx}$. Furthermore, the \emph{mipmap} texture can be used to design,
using a single texture probe, a fast inclusion test of a given cell $c$ at level
$l$ into the shapes: we say that $c$ is in $\Shape$ if its density (retrieved
from the texture probe at level l) is greater than $1/2$. The idea here is to
mimic a kind of adaptive Gauss digitization process.

>>>>

Curvature estimation on discrete or digital surface has also been
widely investigated. In~\cite{CVIU2014}, authors propose digital
versions of Integral Invariant estimators
\cite{Pottmann2007,Pottmann2009} in order to estimate the complete
curvature tensor (mean/Gaussian curvatures, principal curvatures,
principal curvature directions, normal vector field) on digital
surfaces. Such approaches are based on an integration principle using
a ball kernel of a given radius. Additionally, authors have
demonstrated that these estimators have multigrid convergence
properties. In this article, we also propose an efficient GPU
implementation of such estimators to visualize
%, in real-time,
such curvature fields in real-time and
%allowing us
to interactively change the
%explore different
value of the kernel radius.

\vspace{0.2cm} \txtblue{In this paper, we first present the
previous works on curvature tensor estimation. Then, we propose an
efficient GPU approach to extract a triangulated isosurface from a
digital object. Finally, we propose a fully-parallel  GPU pipeline to compute
curvature tensor in real-time and present our results.}

@@

\sloppy In our context, we consider digital shapes (any subset $Z$ of $\Z^d$)
and boundaries of digital shapes $\DigBoundary{Z}$. We denote by $\DSh$ the
Gauss digitization of \comJeremy{a shape $\Shape \subset \R^d$ } in a $d-$dimensional grid with grid step $h$,
\emph{i.e.}  $\DSh \EqDef \{ \vz \in \Z^d,\, h\cdot\vz\in\Shape\}$. For such
digitized set $\DigShape\EqDef\DSh$, $\Body{\DigShape}{h}$ is a subset of $\R^d$
corresponding to the union of hypercubes centered at $h\cdot \vz$ for $\vz\in Z$
with edge length $h$. By doing so, both $\partial\Shape$ and
$\partial\Body{Z}{h}$ are topological boundaries of objects lying in the same
space (see Fig.~\ref{fig:notations}-b). Note that the combinatorial digital
boundary $\DigBoundary{Z}$ of $Z$ made of cells in a cellular Cartesian complex
(\emph{pointels}, \emph{linels}, \emph{surfels}, \ldots), can be trivially
embedded into $\R^d$ such that it coincides with $\partial\Body{Z}{h}$.

\begin{figure}[t]{\small
    \begin{center}
      {\begin{overpic}[width=5cm]{figs/integral2D_2}
          \put(42,28){$\times$}
          \put(46,32){$\vx$}
          \put(55,45){$\Ball{R}{\vx}$}
          \put(1,10){$\dS$}
          \put(15,5){$\Shape$}
      \end{overpic}}~~~~~~
      {\begin{overpic}[width=5cm]{figs/notation_2}
          \put(62,0.5){$h \cdot \vz \in \Shape \cap \Ball{R}{\vx}$}
          \put(55,45){$\Ball{R}{\vx}$}
          \put(42,28.5){$\times$}
          \put(45,32.5){$\vx$}
          \put(45,16.5){$h$}
          %\put(88,39){$\Proj(p)$}
          \put(1.5,10){$\partial \Shape$}
          \put(15,7){$\partial \Body{Z}{h}$}
      \end{overpic}}
      \vspace{-0.4cm}
    \end{center}}
    %
    \caption{Integral invariant computation (\emph{left}) and notations
    (\emph{right}) in dimension 2~\cite{CVIU2014}.\label{fig:notations}}
    %
    \vspace{-0.35cm}
\end{figure}

In~\cite{DGCI2013}, \comJeremy{the 2D digital curvature estimator $\CurvH{R}$
and the 3D digital mean curvature estimator $\MeanCurvH{R}$ are based on the
digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ (we denote such
quantity $\AreaC(Y,h)$ in dimension 2):}
%% Redite
%% \Jaco{Given some point $\vx \in \R^d$, the set $Y$ is the digital intersection of $\DigShape$ with the ball of radius $R/h$ and center $\vx/h$.}
%% where $Y$ is the intersection of $\DigShape$ with a ball
%% $\Ball{R/h}{\vx/h}$ with $\MCard(Y)$, the number of digital points of
%% the intersection between a ball $\Ball{R/h}{\vx/h}$ of digital radius
%% $R/h$ and $\DigShape$ (we denote such quantity $\AreaC(Y,h)$ in
%% dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), \emph{digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \vspace{-0.05cm}
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
\comJeremy{Estimators of both 2D curvature and 3D mean curvature on digital
objects with volumetric integral information are proved to be multigrid
convergent~\cite{DGCI2013}.}
%
For short, when the digital object becomes finer and finer, \ie when the
digitization step $h$ tends to zero, the estimated quantity on
$\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
\txtblue{ associated one on $\Bd{\Shape}$ in $O\left(h^{\frac{1}{3}}\right)$ for
convex shapes with at least $C^3$-boundary and bounded curvature (setting
$R\EqDef kh^\frac{1}{3}$ for some $k\in\R$).}
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
%Setting $R = kh^\frac{1}{3}$ on convex shapes
%with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
%related to the maximal curvature of the shape, they have:
%%
%\begin{align}
%\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
%\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
%&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
%&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
%\end{align}

In~\cite{CVIU2014}, authors have also defined 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  \emph{3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of \comJeremy{$\Ball{R/h}{\vx/h} \cap Z$.}
  % the intersection between the digital shape
  % $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  % $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$, $1$
and $2$ (see Equation~20 of~\cite{CVIU2014} for more details).
% ($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
% $\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
% $\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
% Equation~20 of \cite{CVIU2014} for more details).
\txtblue{
These estimators are proven convergent in $O\left(h^{\frac{1}{3}}\right)$ when setting the ball
radius $h$ in $R = kh^\frac{1}{3}$, where $k$ is a constant related to the
maximal curvature of the shape, on convex shapes with at least $C^3$-boundary
and bounded curvature~\cite{CVIU2014}.}
%More formally, it leads to:
%%
%\begin{align}
%  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
%  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
%  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
%  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
%\end{align}

Additionally, eigenvectors associated to $\hat{\lambda}_1$ and
$\hat{\lambda}_2$ of the covariance matrix are principal
curvature direction estimators $\PrincDirH{1}{R}$ and
$\PrincDirH{2}{R}$. The smallest eigenvector corresponds to the normal
direction $\NormalDirH{R}$ at $\vx$. Convergence results can be found
in~\cite{ChapterIICurvature}.
%%% Changer avec le chapitre de Jaco à un moment

It has been shown that the radius of the ball depends on the geometry of the
underlying shape. In~\cite{DGCI2014}, a proposal was made for a parameter-free
estimation of the radius of the ball by analyzing the shape \wrt the local shape
geometry using maximal digital straight segments of the digital boundary.
In~\cite{SMI2015}, these estimators have been analyzed in scale-space (for a
range of radii) for a given digital shape. This allows to detect features of
the shape thanks to the behavior of estimators on singularities.

As a consequence, for all these integral invariant based approaches, we need to
consider different ball radius which could be time consuming when implemented on
CPU\@. We propose here a fully parallel implementation on GPU allowing us to
change the radius and thus update the estimated quantities in real-time.

@@

Using this spatial data structure, a triangulated mesh can be constructed
%from a Marching Cubes approach \cite{lorensen1987marching} (MC for short): the
using Marching Cubes~\cite{lorensen1987marching} (MC for short): the
triangulation is generated from local triangle patches computed on
local cell configurations.
%Since triangle patches are locally computed,
Such approach is fully parallel and easy to implement on the GPU\@.
However, since adjacent cells may not have the same depth
in the octree, original \textsc{Lorensen} and \textsc{Cline}'s rules need to be
updated (see Figure~\ref{fig_quadtree_partitionning}-\emph{right}).
Many authors have addressed this problem both for primal and
dual meshes
\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.

@@

As illustrated in Figure~\ref{fig_quadtree_partitionning}-\emph{middle}, we
%propose a viewpoint dependent criterion to decide if a cell needs to be refined:
propose a viewpoint dependent criterion to decide if a cell needs to be refined:
the closer we are to the camera, the finer the cells are.
%In addition to the
%distance criterion, we define an angular criterion to prevent refining cells
%outside the camera frustrum.
Such a criterion is also well suited to GPU since it can be evaluated
independently on every cell.
%Hence, we can efficiently determine if a cell must be subdivided or not.
Figure~\ref{fig_lod_octree} illustrates our level of
details (LoD for short) criterion.
% we use. More precisely,
In dimension 2, if $\alpha$ denotes the viewing angle, an object at a distance $d$ from the camera
%with diameter greater than $2\cdot d\cdot\tan(\alpha)$ is projected into more
has a projected size on screen of $2\cdot d\cdot\tan(\alpha)$.
%than one pixel
(see Figure~\ref{fig_lod_octree}-\emph{left}).
%Hence, the
Our distance criterion is based on the ratio (\emph{visibility ratio} in the
following) between the cell diameter $l(c)$ (power of 2 depending on the depth),
and its projected size. %$2\cdot d_c\cdot\tan(\alpha)$, $d_c$ being the distance between the cell
%center and the camera.
For a given cell $c$, split and merge decision are based
on this visibility ratio:
%
\begin{itemize}
  \item $c$ is split if its children
cells have a visibility ratio greater than  constant $k$\,;
  \item $c$ and its sibling cells are merged if their parent cell $c'$ has a visibility ratio lower than $k$\;
  \item otherwise, the cell $c$ stays for the next frame.
\end{itemize}
%
\begin{figure}[!h]
  \begin{center}
    \begin{overpic}[width=3.5cm]{figs/criterion}
      \put(33,40){$\alpha$}
      \put(50,45){$d_c$}
      \put(95,65){$l(c)$}
    \end{overpic}~~~~~
    \includegraphics[width=3cm]{viewlod2_small}~~~~~
    %\includegraphics[width=3cm]{subdivision}
  \end{center}
  \caption{Notations \emph{(left)} and adaptive meshing in dimension 2 using the LoD distance and angular criterion \emph{(right)}.
  %Based on LoD criterion evaluation, the cell buffer is updated between each frame \emph{(right)}.
  }\label{fig_lod_octree}
\end{figure}
%
%In addition to the distance criterion, we add an angular term in order
%to prevent subdivision of cells far from the camera frustrum (see
%Figure~\ref{fig_lod_octree}-\emph{middle}). We do not go into the
%details but please note that stay, split and merge decisions of all cells are computed in
%parallel from all cell morton codes.
Using such a criterion, split and merge decisions are computed in
%of all cells are computed in
parallel from the morton codes of all cells.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{figs/pipeline2}
  \caption{GPU pipeline summary. Data buffers are
    represented in green and computations in red. Each computation
    retrieves data from a buffer and fills a new one.}\label{fig_pipeline}
\end{figure}

@@

\paragraph{Naive Approach.}
\txtblue{A first simple solution  consists in scanning all digital
points, at a given resolution, lying inside the integration domain}
(Fig.~\ref{fig:approx}-$left$) and then \txtblue{estimating} the
geometrical moment as the sum of the geometrical moments of each elementary
cubes lying in the intersection (see
Eq.~(\ref{eq:dig-curvature-estimator-k}) to (\ref{eq:dig-curvature-estimator-k2})).
\txtblue{
This is exactly similar to what is done on the CPU\@. On  the GPU, we can
exploit \emph{mipmap} textures to obtain  multi-resolution information.}
If the input binary object is stored in a
3D texture, GPU hardware constructs a multi-resolution pyramid (mipmap) such that 8
neighboring voxel intensities at level $l$ are averaged to define the voxel
value at level $l+1$. If the level 0 corresponds to the binary input object, at
\txtblue{
a given level $l$, a texture probe at  a point $(x,y,z)$ returns the fraction of $\DSh$ belonging}
to the cube of center $(x,y,z)$ and edge length $2^l$. As a consequence, we can
approximate the volume of $\Ball{R}{\vx}\cap\Shape$ by considering mipmap values
at a given resolution $l$ (Fig.~\ref{fig:approx}-$right$). In this case, errors only
occurs for cells lying inside $\Shape$ (with density 1) not entirely covered by
$\Ball{R}{\vx}$. Furthermore, the \emph{mipmap} texture can be used to design,
using a single texture probe, a fast inclusion test of a given cell $c$ at level
$l$ into the shapes: we say that $c$ is in $\Shape$ if its density (retrieved
from the texture probe at level l) is greater than $1/2$. The idea here is to
mimic a kind of adaptive Gauss digitization process.

\begin{figure}
  \vspace{-0.35cm}
  \begin{center}
    {\includegraphics[width=4.0cm]{figs/approx-naive}}
    {\includegraphics[width=4.0cm]{figs/approx-hierarchical}}
    {\includegraphics[width=4.0cm]{figs/approx-dynamic}}
  \end{center}
  \vspace{-0.35cm}
  \caption{\txtblue{
  	Here are the three approaches we used to evaluate the integrand.
  	(\emph{left}) is the naive approach, where we sum all the digital points that are inside the integrand domain
  	(\emph{middle}) shows the hierarchical decomposition of the ball, thus limiting the number of texture probes to do.
  	(\emph{right}) illustrates the dynamic approach, we probe the textures at a higher resolution and we refine it at each frame until we reach the finer level.
    }}\label{fig:approx}
  \vspace{-0.35cm}
\end{figure}
